Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-02-26 paper.tex
Mass rearranging, hopefully better plot organization

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amssymb,amsmath}
\usepackage[usenames,dvipsnames]{color}
\usepackage[colorlinks,citecolor=RedViolet,urlcolor=blue]{hyperref}
\usepackage{doi}
\usepackage{setspace}
\usepackage[paperwidth=8.5in, paperheight=11in,top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[compact]{titlesec}
\usepackage{abstract}
\usepackage[position=t,singlelinecheck=off,justification=raggedleft]{subcaption}
\usepackage[authoryear,square]{natbib}
\pagestyle{fancy}
\fancyhead[R]{Veibell \thepage}
\fancyhead[L]{}

@@

@@

\section{Introduction}

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period and 1991 and found that the average ion mass density, $M=\rho/n_e$, correlated with geomagnetic activity, with lower $D_{st}$ values corresponding to slightly higher average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$ were compared to $M$.  Most measurements occurred in the MLT range of 12:00 and 18:00 in the plasma trough, and CRRES had a nearly equatorial orbit.

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$. The mass density $\rho$ was estimated using the the Alfven wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was estimated from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincide with with significant changes in $\rho_{eq}$ at an L-shell of 6.8~$R_E$. For five storms over a 20-day period, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$, indicating the possibility of a solar wind control over $\rho_{eq}$.

@@

\cite{Hamilton1988} ... \cite{Gallagher1988} ....

@@

In this work, events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding events, all fill values were replaced with linearly interpolated values.

@@

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold less than $-80$nT results in Figure \ref{Dspec}. A spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 211 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80$~nT that have existing mass density data around onset, so an analysis of this particular combination of factors can't be made.

\subsection{Change in $\rho_{eq}$}
Instead of looking for events based on a threshold of $\rho_{eq}$, we instead looked for a certain amount of change in $\rho_{eq}$ as the basis for event onset. Both raw change and percent change were done in case of bias towards high or low $\rho_{eq}$ periods, respectively, but neither showed any correlation to a change in $D_{st}$.

@@

Looking at GOES 2,5,6, and 7 (Figures \ref{GOES67,GOES25}) shows that this post-onset $D_{st}$ spike happens in almost every case for periods of high $F_{10.7}$.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{paperfigures/F107MDAllData.eps}
\includegraphics[scale=0.40]{paperfigures/ccplot.eps}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $\log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{1pt}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT }
\label{Dspec}
\end{figure}

@@

@@

\section{Appendix}
\hfill
\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst30.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst30-valid.eps}
\caption{$D_{st} < -30$ nT events, $\rho_{eq}$ binned by $F_{10.7}$. Lower panel: valid hourly points}
\end{figure}
\hfill
\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-rhoeq40.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-rhoeq40-valid.eps}
\caption{$\rho_{eq} > 40$ amu/cm$^3$ events, $\rho_{eq}$ binned by $F_{10.7}$}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/DstRhoThresh-1983-1991.eps}
\includegraphics[scale=0.45]{paperfigures/DstRhoThresh-1987-1991.eps}
\caption{High $F_{10.7}$ events - Low $F_{10.7}$ events, both with baselines removed. 83-91 and 87-91}
\end{figure}

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/DstRhoThresh-1983-1987.eps}
\includegraphics[scale=0.45]{paperfigures/DstRhoThresh-1981-1983.eps}
\includegraphics[scale=0.45]{paperfigures/RandRhoThresh-1983-1991.eps}
\caption{High $F_{10.7}$ events - Low $F_{10.7}$ events, both with baselines removed. 83-87 and 81-83. Bottom: Random "events" from 1983-1991}
\end{figure}
\clearpage

>>>>

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amssymb,amsmath}
\usepackage[usenames,dvipsnames]{color}
\usepackage[colorlinks,citecolor=RedViolet,urlcolor=blue]{hyperref}
\usepackage{doi}
\usepackage{setspace}
\usepackage[paperwidth=8.5in, paperheight=11in,top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[compact]{titlesec}
\usepackage{abstract}
\usepackage[position=t,singlelinecheck=off,justification=raggedleft]{subcaption}
\usepackage[authoryear,square]{natbib}
\usepackage{aas_macros}
\pagestyle{fancy}
\fancyhead[R]{Veibell \thepage}
\fancyhead[L]{}

\newcommand{\vinote}[1]{\textcolor{red}{\textbf{#1}}} %If you want \inote{#} instead of just \note #, for inline notes
\def\vnote#1\par{\textcolor{red}{\textbf{#1}}\\} %Make note bold and red, re-add newline
\newcommand{\req}{\ensuremath{\rho_{eq}}}

@@

\vnote More detail

@@

\section{Introduction}
\cite{Gallagher1988EmpiricalModelPlasmasphere} \vinote{Too far back?} first models the the plasmasphere as a simple linear fit dependent on L shell, geographic height, geomagnetic latitude, and $K_p$. \cite{Hamilton1988RingCurrentDevelopment} simultaneously analyzes a strong geomagnetic storm by comparing empirical energy measurements of the ring current to $D_{st}$, along with measuring ring current compositions before, during, and after the storm.

\cite{Lotoaniu1999PlasmaMassDensity} uses CRRES to compare in-situ measured plasma mass density to that inferred by ground-based ULF wave measurements extended to the plasmatrough.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 and found that the average ion mass, $M=\rho/n_e$, correlated with geomagnetic activity, with more negative $D_{st}$ values corresponding to slightly higher average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$, with an increasing correlation as $K_p$ was smoothed to longer timescales.  The CRRES measurements used occurred in the MLT range of 12:00 and 18:00 mostly in the plasma trough (between 2-7$R_E$) on an elliptical, nearly equatorial orbit.

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfven wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

\cite{Takahashi2010} \vinote{Necessary repetition?} noted that downward spikes in the $D_{st}$ index coincide with with significant changes in $\rho_{eq}$ at an L-shell of 6.8~$R_E$. For five storms over a 20-day period, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$, indicating the possibility of a solar wind control over $\rho_{eq}$.

@@

@@

In this work, events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values.

@@

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold less than $-80$nT finds both result in a spike in mass density, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 211 events.

\subsection{Change in $\rho_{eq}$}
Instead of looking for events based on a threshold of $D_{st}$, we instead looked for a certain amount of change in $\rho_{eq}$ as the basis for event onset. Both raw change and percent change were done in case of bias towards high or low $\rho_{eq}$ periods, respectively, but neither showed any correlation to a change in $D_{st}$.

@@

Looking at GOES 2,5,6, and 7 shows that this post-onset $D_{st}$ spike happens in almost every case for periods of high $F_{10.7}$. Creating a linear model directly relating variables (and combinations of variables) to $\rho_{eq}$ also shows that $F_{10.7}$ most often has the highest correlation with $\rho_{eq}$. A sample of these correlations is shown in Table \ref{perltable}. The correlations were made by training a one-input linear model of 5-hour post-onset medians of each variable (or variables) against the median of $\rho_{eq}$ at, and four hours after, event onset. This was done by randomly sampling half of the data, generating the model, and testing against the other half, doing this process 100 times and then taking the median correlations.

\input{"tables/perltable.tex"}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{paperfigures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{paperfigures/ccplot.eps}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $\log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

@@

@@

\section{Appendix}
\hfill

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{1pt}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT }
\label{Dspec}
\end{figure}

@@

\hfill
\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst30.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst30-valid.eps}
\caption{$D_{st} < -30$ nT events, $\rho_{eq}$ binned by $F_{10.7}$. Lower panel: valid hourly points}
\end{figure}
\hfill
\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-rhoeq40.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-rhoeq40-valid.eps}
\caption{$\rho_{eq} > 40$ amu/cm$^3$ events, $\rho_{eq}$ binned by $F_{10.7}$}
\end{figure}
\clearpage

@@
